Two Decades of Smalltalk VM Development

### 2018-08-17 paper.tex
Comprehensive pass over the English in the paper, eliminating unnecessary commas, indefinite articles, etc. Added an anecdote as a footnote describing why the simulator was retained for JIT development. Moved description of Lemming Debugging to its own section (its cool!). Described going in the pposite direction to the in-image compilation facade. Fixed capitalisation in a citation.

\openSmalltalkVM is a virtual machine (VM) for languages in the Smalltalk family (e.g. Squeak, Pharo) which is itself written in a subset of Smalltalk that can easily be translated to C.  Development is done in Smalltalk, an activity we call ``Simulation''. The production VM is derived by translating the core VM code to C. As a result, two execution models coexist: simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler.

%\openSmalltalkVM is a VM for languages in the Smalltalk family which is itself written in a subset of Smalltalk that can easily be translated to C.  Development is done in Smalltalk, an activity we call "Simulation", and the support code for which is "The Simulator". The production VM is derived by translating the core VM code to C, combining the generated C code with a set of platform-specific support files, and compiling with the platform's C compiler. Two execution models are effectively available, simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler. Simulation is used to develop and debug the VM. Production is used to release the VM.

@@

To specify the virtual machine (VM), the Smalltalk-80 team at Xerox PARC wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}. In 1995, members of the same team built Squeak \cite{SqueakByExample}, an open-source Smalltalk dialect, and its VM \cite{BackToTheFuture}, written in Smalltalk using the code from Smalltalk-80 \cite{blueBook} as a starting point. Part of the code base was, however, narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation.  Development was done in Smalltalk, an activity we call "Simulation", and the support code for which is "The Simulator". The production VM is derived by translating the core VM code to C, combining this generated C code with a set of platform-specific support files, and compiling with the platform's C compiler.

@@

Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for various Smalltalk-like systems such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and Newspeak \cite{NewspeakOopsla}. 
As the VM evolved, the original simulator co-evolved as a tool to develop and debug the VM. The metadata used to guide the translation process was replaced by \emph{pragmas} \cite{Pragmas}.  
%One change moved the meta data that guides translation out of the executable Smalltalk into pragmas \cite{Pragmas}.  
But the most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT) \cite{CogJIT}. 
%The code base was liberally sprinkled with assertions. 
The spaghetti stack was mapped to a more conventional stack frame organisation in a stack zone of a few hundred kilobytes organised into small pages, a scheme called context-to-stack mapping \cite{EfficientST80,StackMappingVW}.  Machine code generated by the JIT was executed by binding multiple processor simulators (Bochs \cite{Bochs} for x86 and x64, SkyEye for ARMv6, Smalltalk code for 32-bit MIPS).  
Finally, a new object representation and garbage collector was added to improve performance and also to support 64-bits \cite{Forwarders}.  This required refactoring the interpreter to allow the object representation to be chosen at start-up\footnote{The JIT was written with this eventuality in mind.}. %Bochs was used a second time for x64 support.

One of the key design decisions in \openSmalltalkVM was to keep the core components (i.e. the interpreter, the JIT and the memory manager) written in Slang and not in C/C++.  By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, it is possible to simulate the whole VM execution. This allows development and debugging of the VM with the Smalltalk development tools resulting in a live programming experience for VM development.

@@

The VM executable is generated in a two step process. Firstly, the Slang-to-C compiler translates the Slang code (interpreter, JIT and GC code) to C code, generating a few C files. This first step usually takes several seconds. Secondly, the C compiler, depending on the platform, LLVM, GCC or MSVC, translates all the C files (generated C files and platform-specific C files) into an executable. This second step can take up to a few minutes the first time, but usually only takes a few seconds, since unchanged C files are kept in the C compiler cache.

@@

\paragraph{Snapshots.}
Smalltalk is an object \emph{system}, rather than a language. The entire system, including its development tools and application code is stored in a snapshot file, which is essentially a memory dump of the entire heap. 
%A Smalltalk system is started by loading a snapshot file into a VM. The snapshot includes objects such as the classes, the compiled methods in the form of bytecodes and the running processes. At start-up, the VM restores the state of all objects in memory, swizzling pointers as required, and resumes execution in the process active when the snapshot was created.  
%A new snapshot can be made during or at the end of the session.
When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development. 
More precisely, the snapshot includes objects (such as the classes), the compiled methods in the form of bytecodes and the running processes.
Developing applications consists essentially in writing and editing code, which effectively installs, modifies and removes classes and compiled methods from objects. Programming may be done live, as the application under development is running. %, for example creating or revising methods in the debugger, and continuing. 
For example, objects may have their shape changed on the fly as instance variables are added and removed%\footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello \cite{Monticello}}.
. A new snapshot can be made during or at the end of the development session.

Snapshots can be used to avoid long start-up times when fixing specific bugs. The VM can be run to a point where the bug is about to be injected and the system can be snapshotted at this point. Then, multiple analyses of the bug can be undertaken by loading the snapshot and resuming execution, either in the normal VM or in the simulator, short cutting the time to reach the bug.

@@

\paragraph{Production memory layout.}Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM at runtime and the simulated counter-part. Let us detail briefly the memory used by the production VM, on the top of the figure. On the left side, which is usually the low addresses, we can first find Text, the memory section holding the native code of the VM (the compiled code of the interpreter, the memory manager and the JIT itself, but not the code compiled by the JIT). Then, Data holds initialized and uninitialized data, mainly the C variables used globally in the VM.

On higher addresses, we can find the beginning of the memory managed by the VM. At start-up, the VM mmaps a memory region used for the executable code generated by the JIT (this optional execution exists only if the JIT is enabled), for the young objects and for the old objects present in the snapshot started plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector. Later during execution, as old space grows, new memory regions are mmap'ed at higher addresses to store other old objects.

Usually at very high addresses, we find the C stack. Room in the C stack is allocated at VM start-up to hold the Smalltalk stack zone. Both stacks are disjoint and managed differently, the stack zone being broken up into small stack pages \cite{StackMappingVW}.

\paragraph{Simulation memory layout.}In the simulator, the heap is stored as a large contiguous byte array. References between objects are indices inside the byte array instead of pointers. All the Slang variables, normally translated to C variables, are simulated as Smalltalk objects. They use specific wrapper classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs, etc.). The Slang code is executed as Smalltalk code. The Smalltalk stack zone is represented in \openSmalltalkVM as a double linked list of stack pages which are maintained by the VM. In the interpreter-only simulator, the stack zone is a Smalltalk Array object. In the full simulator, the stack zone is in the byte array, to allow access to the stack from generated machine code by the processor simulator\footnote{This difference is historical, the interpreter-only was implemented first in the easiest way possible, but it did not work out with the JIT.}.

@@

To be able to reproduce the same bug the exact same multiple times in a row, we designed the simulator to be as deterministic as possible. The most important things to note are the following:
\begin{itemize}
\item Simulated memory is not subject to Address Space Layout Randomisation,
\item A synthetic clock is used so that time advances in lock step with code execution.
\end{itemize}

@@

In addition to the interpreter simulator, simulating the JIT requires to simulate the execution of the native code it generates. The JIT itself is written in Slang and simulated with the Smalltalk execution model. To simulate the machine code, the start of the byte array representing the memory holds, when the JIT is enabled, the machine code generated at runtime.

Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6, Smalltalk code for 32-bit MIPSLE) were implemented so that the machine code can be executed safely. All but the MIPSLE simulators are accessed through %Smalltalk primitives.  These are effectively a combination of an 
FFI calls and surrounding glue code.
% to invoke foreign code plus Smalltalk code that is executed if (the code invoked by) the primitive returns an error. 
The next paragraphs describe in detail how this scheme is used to ensure that simulation and production machine code are as similar as possible and how simulated machine code is interfaced to the rest of the simulation. %, and the debugging facilities for machine code that are available.

@@

A key requirement to enable effective development of the JIT is the code generated during simulation to be as close as possible to the code generated by the production JIT.  The code is of course not bit-identical; the address space is laid out very differently in memory to the simulation's byte array and the simulator is not subject to Address Space Layout Randomisation. But we benefit from the code generated under simulation being otherwise identical to that generated in the production VM and we go to some effort to achieve this.  The technique we use, involving illegal addresses, also allows us to interface with machine code the Smalltalk objects that comprise the Simulation, the interpreter, memory manager and JIT, and variables within them. To explain this let's consider how the interpreter references a stack frame.

The interpreter has framePointer, stackPointer and instructionPointer variables used to interpret code. When machine code wants to enter the runtime, for example to have the runtime look up a send on a cache miss, or to invoke the scavenger when young space is full, it needs to record the current stack frame, writing the native stack and frame pointers into the stackPointer and framePointer variables, and then to call the relevant routine. The runtime can then access the current machine code stack frame though these variables. In the production VM, the stackPointer and framePointer variables are C variables at known addresses in Data, and routines in the runtime have been compiled from C and exit in Text. Referencing these variables and invoking runtime routines from machine code is therefore straight-forward.  But in the simulation they are inaccessible. First of all they are instance variables inside Smalltalk objects, objects that have strong encapsulation, accessible only through messages, and second of all they are alongside the simulation's byte array, not within it, therefore the processor simulator has no access to these variables.

%NOTE: I almost only commented part of what Eliot wrote but did not change anything. I basically removed implementation specific details.
The JIT maintains a set of Dictionaries (hash maps) that map integers representing addresses to closures that get or set variables in the simulation objects, or invoke runtime routines, in both cases by sending messages to the simulation objects. As code is JITted and specific variables and runtime routines are referenced, the JIT manufactures a unique illegal address to reference the variable or routine, and uses this as the key in the relevant map, creating and storing the relevant closure accessor as the value.  When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction\footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it raises a ProcessorException.  %Our processor simulators are invoked through a "primitive" interface, much like an FFI call, and the attempt to raise an exeption is caught by wrapper code that returns an error result from the primitve invocation; in Smalltalk terminology the primitive fails.  The primitive failure code, Smalltalk code in the method that contains the primitive invocation, then analyses the offending instruction and constructs a Smalltalk ProcessorException object.  This Smalltalk exception has a type (\#read, \#write, \#jump, \#call), the address of the memory location being read, written, jumped to or called, a program counter (pc) holding the instruction's address, a next pc, and an accessor for the register, if any, that is the target of the read or the source of the write.  
The processor simulator is invoked in the context of an exception handler for such exceptions which invokes %a Smalltalk method, handleSimulationTrap:, that invokes 
the relevant closure. Once the closure has completed, the processor simulator is advanced to the next pc and simulation resumes. In this way, machine code generated in simulation can access arbitrary simulation objects while remaining essentially identical to the generated machine code in the production VM. Further, ProcessorException, and a handful of other methods implemented by a processor simulator makes the interfacing of machine code to the Smalltalk part of the simulation portable and processor independent.

@@

There is a small polymorphic scheme for breakpoints.  Booleans, integers, and arrays understand isBreakpointFor: pc.  So if the breakPC object is false there is no breakpoint.  If the breakPC is true then any pc is a potential breakpoint.  If it is an integer then that pc is a potential breakpoint.  If the pc is an Array then the pc is a breakpoint if any value in the array answers true to isBreakpointFor:. The scheme could trivially be extended to include intervals. If breakPC isBreakpointFor: pc is true then the breakBlock is sent shouldStopIfAtPC: pc.  booleans and closures understand shouldStopIfAtPC:.  true and false make the breakpoint unconditional or disabled respectively, and a closure evaluates itself, allowing one to specify arbitrarily complex breakpoints.  For example, one can specify that execution should stop at a particular pc if the top two elements on the stack satisfy some criterion.  It is much more powerful and much simpler to use than typical machine-level debuggers, and it can be extended as one is debugging (for example, adding interval support during the middle of a debugging session if required).

@@

Simulating the whole VM requires going through the whole start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface. The whole start-up takes, in the simulator, around 15 seconds on a recent Macbook pro. While developing the JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle, which we want to avoid. In addition, when developing a specific part of the JIT, it is convenient to compile a well-chosen simple method exercising that specific new part. This is difficult to do with the full simulator since that new part of the JIT might be called by the start-up logic, leading to the erratic behavior before expected.

@@

To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory. To implement this we built a facade, which masquerades as the simulation's memory manager, and translates the state of the bytecode compiled method in the current simulation into state that makes it appear as if it were resident in a simulation byte array.  This includes mock addresses for all the objects the JIT may require to generate the machine code of a given method. Figure \ref{fig:InImageCompiler} summarizes the in-image compilation process. Note that this technique applies for the baseline JIT, which translate a single bytecode method into machine code; adaptive optimizations and speculative optimizations are debugged differently.

@@

@@

\paragraph{Decorated Disassembly.}
Each processor simulator is expected to be able to disassemble its machine code.  The Smalltalk code for each processor simulator implements a simple parser that identifies constants and field offsets embedded in machine code and passes these to routines in the JIT and the interpreter that lookup addresses, matching them to objects, in particular message selectors and classes, and to method local variable names, rendering the machine code much more readable than machine code displayed in a typical machine oriented debugger.  For example, Figure \ref{fig:ASMDecorated} is a snippet of decorated disassembly for x64.

This code is the end of the entry sequence that checks the stack pointer against the stack limit for the end of the current stack page, which is also used to break out of machine code to service interrupts or events such as invoking the scavenger, etc.  \emph{\%rbx} is used as a base register pointing at the variables in interpreter.  Following that is the code for \emph{index := self indexOf: anElement startingAt: start} where \emph{anElement} and \emph{start} are method parameters and \emph{index} is a local variable.  HasBytecodePC and IsSendCall are decorations that identify important locations in machine code, including object references, runtime calls, etc. These points are specified using metadata implemented as a simple bytecoded language, a stream of such bytes being added to the end of each machine code method.  This metadata is parsed when the garbage collector needs to locate object references in machine code, when linked sends must be relocated when the machine code zone is compacted to make space by throwing away some number of least recently used methods, and when machine code pcs must be mapped to their corresponding bytecode pcs for debugging purposes. %to preserve the illusion that the VM is a bytecode interpreter using a spaghetti stack of Context objects.

@@

\subsection{Implementing a new full GC algorithm}

@@

In this section we first discuss the VM simulator limitations and how we work around them, then the interaction between the simulator and bootstrapping processes. Lastly, we mention some relevant related works.

@@

%However, so far, most production VMs (Java, Javascript, etc.), even after the huge recent investments in the Javascript VMs by the four major web vendors, are still compiling through the C/C++ compiler and are not metacircular. 
%Hence, although a metacircular VM has interesting advantages, it is not clear it is that convenient to build a VM in such a way.

>>>>

\openSmalltalkVM is a virtual machine (VM) for languages in the Smalltalk family (e.g. Squeak, Pharo) which is itself written in a subset of Smalltalk that can easily be translated to C.  Development is done in Smalltalk, an activity we call ``Simulation''. The production VM is derived by translating the core VM code to C. As a result, two execution models coexist: simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through a C compiler.

%\openSmalltalkVM is a VM for languages in the Smalltalk family which is itself written in a subset of Smalltalk that can easily be translated to C.  Development is done in Smalltalk, an activity we call "Simulation", and the support code for which is "The Simulator". The production VM is derived by translating the core VM code to C, combining the generated C code with a set of platform-specific support files, and compiling with the platform's C compiler. Two execution models are effectively available, simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through a C compiler. Simulation is used to develop and debug the VM. Production is used to release the VM.

@@

To specify their virtual machine (VM), the Smalltalk-80 team at Xerox PARC wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}. In 1995, members of the same team built Squeak \cite{SqueakByExample}, an open-source Smalltalk dialect, and its VM \cite{BackToTheFuture}, written in Smalltalk using the code from Smalltalk-80 \cite{blueBook} as a starting point. Part of the code base was, however, narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation.  Development was done in Smalltalk, an activity we call "Simulation", and the support code for which is "The Simulator". The production VM is derived by translating the core VM code to C, combining this generated C code with a set of platform-specific support files, and compiling with the platform's C compiler.

@@

Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for various Smalltalk-like systems such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and Newspeak \cite{NewspeakOopsla}. 
As the VM evolved, the original simulator co-evolved as a tool to develop and debug the VM. The metadata used to guide the translation process was replaced by \emph{pragmas} \cite{Pragmas}.  
%One change moved the meta data that guides translation out of the executable Smalltalk into pragmas \cite{Pragmas}.  
But the most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT) \cite{CogJIT}. 
%The code base was liberally sprinkled with assertions. 
The spaghetti stack was mapped to a more conventional stack frame organisation in a stack zone of a few hundred kilobytes organised into small pages, a scheme called context-to-stack mapping \cite{EfficientST80,StackMappingVW}.  A JIT was written. Machine code generated by the JIT was executed by binding multiple processor simulators (Bochs \cite{Bochs} for x86 and x64, SkyEye for ARMv6, Smalltalk code for 32-bit MIPS).  
Support for multiple bytecode sets was added, initially to support Newspeak, and more recently to support adaptive optimization.
Finally, a new object representation and garbage collector was added to improve performance and also to support 64 bits \cite{Forwarders}.  This required refactoring the interpreter to allow the object representation to be chosen at start-up\footnote{The JIT was written with this eventuality in mind.}. %Bochs was used a second time for x64 support.  Slang was also extended with type inference to allow Smalltalk code that is independent of word size by virtue of Smalltalk's infinite precision arithmetic to be used in both 32 and 64 bit contexts.

One of the key design decisions in \openSmalltalkVM was to keep the core components (i.e. the interpreter, the JIT and the memory manager) written in Slang and not in C/C++.  By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, it is possible to simulate the whole VM execution. This allows development and debugging of the VM with the Smalltalk development tools resulting in a live programming experience for VM development.\footnote{Arriving at the startup that first funded the OpenSmalltalk-VM, the first author had considerable experience maintaining the 2nd generation Deutsch Schiffmann VM written in C, and was sceptical that the Squeak Simulator would be viable for implementing a JIT.  The startup wanted incremental development so as to reap value early and reduce risk, so it asked first for an interpreter using context-to-stack mapping. Writing this initial deliverable within the simulator was so much more productive and pleasurable than his previous work in C that he happily decided to stay with the simulator and invest in the necessary machine code simulation infrastructure.}

@@

The VM executable is generated in a two step process. Firstly, the Slang-to-C compiler translates the Slang code (interpreter, JIT and GC code) to C code, generating a few C files. This first step usually takes several seconds. Secondly, the C compiler, depending on the platform, LLVM, GCC or MSVC, translates all the C files (generated C files and platform-specific C files) into an executable. This second step can take up to a few minutes the first time, but usually only takes a few seconds, since object files for unchanged C files do not need to be recreated.

@@

\paragraph{Snapshots.}
Smalltalk is an object \emph{system}, rather than a language. The entire system, including its development tools and application code is stored in a snapshot file, which is essentially a memory dump of the entire heap. 
%A Smalltalk system is started by loading a snapshot file into a VM. The snapshot includes objects such as the classes, the compiled methods in the form of bytecodes and the running processes. At start-up, the VM restores the state of all objects in memory, swizzling pointers as required, and resumes execution in the process active when the snapshot was created.  
%A new snapshot can be made during or at the end of the session.
When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development. 
More precisely, the snapshot includes objects (such as the classes), the compiled methods in the form of bytecodes and the running processes.
Developing applications consists essentially in writing and editing code, which installs, modifies and removes classes and compiled methods from objects. Programming may be done live, as the application under development is running. %, for example creating or revising methods in the debugger, and continuing. 
For example, objects may have their shape changed on the fly as instance variables are added and removed%\footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello \cite{Monticello}}.
. A new snapshot can be made during or at the end of the development session.

Snapshots can be used to avoid long start-up times when fixing specific bugs. The VM can be run to a point where the bug is about to be encountered and the system can be snapshotted at this point. Then multiple analyses of the bug can be undertaken by loading the snapshot and resuming execution, either in the normal VM or in the simulator, short cutting the time to reach the bug.

@@

\paragraph{Production memory layout.}Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM in the prducton VM and its simulated counter-part. Let us detail briefly the memory used by the production VM, on the top of the figure. On the left side, which is usually the low addresses, we can first find Text, the memory section holding the native code of the VM (the compiled code of the interpreter, the memory manager and the JIT itself, but not the code compiled by the JIT). Then, Data holds initialized and uninitialized data, mainly the C variables used globally in the VM.

At higher addresses we find the beginning of the memory managed by the VM. At start-up the VM mmaps a memory region used for the executable code generated by the JIT (this optional execution exists only if the JIT is enabled), for new space (initially empty) and for the old objects present in the snapshot plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector. Later during execution, as old space grows, new memory regions are mmap'ed at higher addresses to store other old objects.

Usually at a very high addresse we find the C stack. Room in the C stack is allocated at VM start-up to hold the Smalltalk stack zone. Both stacks are disjoint and managed differently, the stack zone being broken up into small stack pages \cite{StackMappingVW}.

\paragraph{Simulation memory layout.}In the simulator, the heap is stored as a large contiguous byte array. References between objects are indices inside the byte array instead of pointers. All the Slang variables, normally translated to C variables, are simulated as Smalltalk objects. They use specific wrapper classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs, etc.). The Slang code is executed as Smalltalk code. The Smalltalk stack zone is represented in \openSmalltalkVM as a doubly linked cycle of stack pages which are maintained by the VM. In the interpreter-only simulator, the stack zone is a Smalltalk Array object. In the full simulator, the stack zone is in the byte array, to allow access to the stack from generated machine code by the processor simulator\footnote{This difference is historical, the interpreter-only version was implemented first in the easiest way possible, but out with the JIT the Smalltalk stack must be accessible to the processor simulator.}.

@@

To be able to reproduce the same bug exactly multiple times in a row, we designed the simulator to be as deterministic as possible. The most important features are:
\begin{itemize}
\item Simulated memory is not subject to Address Space Layout Randomisation,
\item A synthetic clock is used so that time advances in lock step with code execution.
\end{itemize}

@@

In addition to the interpreter simulator, simulating the JIT requires simulating execution of the native code it generates. The JIT itself is written in Slang and simulated with the Smalltalk execution model. When the JIT is enabled, the start of the byte array representing the memory is used to hold the machine code it generates at runtime.

Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6, Smalltalk code for 32-bit MIPSLE) were implemented so that the machine code can be executed safely. All but the MIPSLE simulators are accessed through %Smalltalk primitives.  These are effectively a combination of an 
FFI calls and surrounding glue code.
Each processor simulator is invoked via calls that supply the byte array as a parameter, and effectively the byte array is the processor simulator's memory, while the processor simulator contains its register state which is accessed via Smalltalk's support for external memory access. \footnote{http://www.squeaksource.com/Alien}
% to invoke foreign code plus Smalltalk code that is executed if (the code invoked by) the primitive returns an error. 
The next paragraphs describe in detail how this scheme is used to ensure that simulation and production machine code are as similar as possible and how simulated machine code is interfaced to the rest of the simulation. %, and the debugging facilities for machine code that are available.

@@

A key requirement to enable effective development of the JIT is that the code generated during simulation be as close as possible to the code generated by the production JIT.  The code is of course not bit-identical; the address space is laid out very differently in memory to the simulation's byte array and the simulator is not subject to Address Space Layout Randomisation. But we benefit from the code generated under simulation being otherwise identical to that generated in the production VM and we go to some effort to achieve this.  The technique we use, involving illegal addresses, also allows us to interface generated machine code with the Smalltalk objects that comprise the Simulation, the interpreter, memory manager and JIT, and variables within them. To explain this let's consider how the interpreter references a stack frame.

The interpreter has \emph{framePointer}, \emph{stackPointer} and \emph{instructionPointer} variables used to interpret code. When machine code wants to enter the runtime, for example to have the runtime look up a send on a cache miss, or to invoke the scavenger when young space is full, it needs to record the current stack frame, writing the native stack and frame pointers into the \emph{stackPointer} and \emph{framePointer} variables, and then to call the relevant routine. The runtime can then access the current machine code stack frame though these variables. In the production VM, the \emph{stackPointer} and \emph{framePointer} are C variables in Data, and routines in the runtime have been compiled from C and exist in Text, and all have fixed addresses. Referencing these variables and invoking these runtime routines from machine code is therefore straight-forward.  But in the simulation they are inaccessible. First of all they are instance variables inside Smalltalk objects, objects that have strong encapsulation, accessible only through messages, and second of all they are alongside the simulation's byte array, not within it, therefore the processor simulator has no access to these variables.

%NOTE: I almost only commented part of what Eliot wrote but did not change anything. I basically removed implementation specific details.
The JIT maintains a set of Dictionaries (hash maps) that map integers representing addresses to closures that get or set variables in the simulation objects, or invoke runtime routines, in both cases by sending messages to the simulation objects. As code is JITted and specific variables and runtime routines are referenced, the JIT manufactures a unique illegal address to reference each variable or routine, and uses this as the key in the relevant map, creating and storing a suitable closure accessor as the value.  When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction\footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it raises a ProcessorException.  %Our processor simulators are invoked through a "primitive" interface, much like an FFI call, and the attempt to raise an exeption is caught by wrapper code that returns an error result from the primitve invocation; in Smalltalk terminology the primitive fails.  The primitive failure code, Smalltalk code in the method that contains the primitive invocation, then analyses the offending instruction and constructs a Smalltalk ProcessorException object.  This Smalltalk exception has a type (\#read, \#write, \#jump, \#call), the address of the memory location being read, written, jumped to or called, a program counter (pc) holding the instruction's address, a next pc, and an accessor for the register, if any, that is the target of the read or the source of the write.  
The processor simulator is always invoked in the context of an exception handler for such exceptions which invokes %a Smalltalk method, \emph{handleSimulationTrap:}, that invokes 
the relevant closure. Once the closure has completed, the processor simulator is advanced to the next pc and simulation resumes. In this way, machine code generated in simulation can access arbitrary simulation objects while remaining essentially identical to the generated machine code in the production VM. Further, ProcessorException, and a handful of other methods implemented by a processor simulator makes the interfacing of machine code to the Smalltalk part of the simulation processor independent.

@@

There is a small polymorphic scheme for breakpoints.  Booleans, integers, and arrays understand \emph{isBreakpointFor: pc}.  So if the \emph{breakPC} variable is \emph{false} there is no breakpoint.  If \emph{breakPC} is \emph{true} then any pc is a potential breakpoint.  If it is an integer then that pc is a potential breakpoint.  If the pc is an \emph{Array} then the pc is a breakpoint if any value in the array answers true to \emph{isBreakpointFor:}. The scheme could trivially be extended to include intervals. If \emph{breakPC isBreakpointFor: pc} is \emph{true} then the \emph{breakBlock} is sent \emph{shouldStopIfAtPC: pc}.  booleans and closures understand \emph{shouldStopIfAtPC:}.  \emph{true} and \emph{false} make the breakpoint unconditional or disabled respectively, and a closure evaluates itself, allowing one to specify arbitrarily complex breakpoints.  For example, one can specify that execution should stop at a particular pc if the top two elements on the stack satisfy some criterion.  It is much more powerful and much simpler to use than typical machine-level debuggers, closures can be created at any point in any tool with a text interface, such as the debugger, and the scheme can be extended as one is debugging (for example, adding interval support during the middle of a debugging session if required).

@@

Simulating the whole VM requires going through the entire Smalltalk system start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface. In the simulator start-up takes around 15 seconds on a recent MacBook Pro. While developing the JIT, this start-up time may still be too long and degrade the live programming experience to an edit-compile-run cycle, which we prefer to avoid. In addition, when developing a specific part of the JIT, it is convenient to compile a well-chosen simple method exercising that specific new part. This is difficult to do with the full simulator since that new part of the JIT might be called by the start-up logic, leading to the erratic behavior before expected.

@@

To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as \emph{true}, \emph{false} or \emph{nil}) as if they were in the simulated memory. To implement this we built a facade, which masquerades as the simulation's memory manager, and translates the state of the bytecode compiled method in the current simulation into state that makes it appear as if it were resident in a simulation byte array.  This includes mock addresses for all the objects the JIT may require to generate the machine code of a given method. Figure \ref{fig:InImageCompiler} summarizes the in-image compilation process. Note that this technique applies for the baseline JIT, which translate a single bytecode method into machine code; adaptive optimizations and speculative optimizations are debugged differently.

@@

We also go in the opposite direction to the facade.  Smalltalk provides a pretty printer for bytecoded methods.  The class \emph{VMCompiledMethodProxy} is polymorphic with \emph{CompiledMethod} and makes methods in the simulation's heap appear to be ordinary \emph{CompiledMethods} in the Smalltalk system, allowing the pretty printer, and other tools, to be applied to methods in the simulation.  This scheme can be extended as required; potentially any object in the simulation can be reified to be inspected using the Smalltalk tool suite.

@@

\paragraph{Decorated Disassembly.}
Each processor simulator is expected to be able to disassemble its machine code.  The Smalltalk code for each processor simulator implements a simple parser that identifies constants and field offsets embedded in that disassembly and passes these to routines in the JIT and the interpreter that lookup addresses, matching them to objects, in particular message selectors and classes, and to method local variable names, rendering the disassembly much more readable than that displayed in a typical machine oriented debugger.  For example, Figure \ref{fig:ASMDecorated} is a snippet of decorated disassembly for x64.

This code is the end of the entry sequence that checks the stack pointer against the stack limit for the end of the current stack page, which is also used to break out of machine code to service interrupts or events such as invoking the scavenger, etc.  \emph{\%rbx} is used as a base register pointing at the variables in interpreter.  Following that is the code for \emph{index := self indexOf: anElement startingAt: start} where \emph{anElement} and \emph{start} are method parameters and \emph{index} is a local variable.  HasBytecodePC and IsSendCall are decorations that identify important locations in machine code, including object references, runtime calls, etc. These points are specified using metadata implemented as a simple bytecoded language, a stream of such bytes being added to the end of each machine code method.  This metadata is parsed when the garbage collector needs to locate object references in machine code, when linked sends must be relocated when the machine code zone is compacted to make space by throwing away some number of least recently used methods, and when machine code pcs must be mapped to their corresponding bytecode pcs since at the Smalltalk level only bytecode pcs are seen. %to preserve the illusion that the VM is a bytecode interpreter using a spaghetti stack of Context objects.

@@

\subsection{Initial JIT Development and New Back Ends}
When the JIT was first being developed and was at a state where bytecoded methods could actually be translated to machine code, development fell into a very productive and enjoyable pattern.  As a bytecode was encountered for which the template had yet to be written execution would stop in the debugger, which would create a skeleton implementation for the template method from the MessageNotUnderstood exception. The programmer would implement the body of the method and execution would continue.  Were the new template definition to be incorrect and cause a bug during JIT generation, rather than when the generated machine code was run, then the debugger could be used to wind execution back to the beginning of the JITting of the current method and resume execution, repeating the cycle until disassembly looked good, somewhat like lemming debugging above.  As a new method was encountered and JITted yet more templates would be encountered.  Similarly as the abstract machine instruction to concrete machine instruction mappings were added these could be implemented in the debugger, an experience repeated when new back ends were added. Much of the JIT was written in this interactive live style.  This style encourages tool creation because one can create, extend and polish tools in the middle of debugging sessions, enhancing them to make sense of ones current predicament.  Immediate feedback makes tool investment cheap and cheerful.

\subsection{Debugging New GC Algorithms and Lemming Debugging}

Debugging graph rewriting garbage collectors can be hard; bugs leave the heap in a scrambled state, and figuring out what happened when one is left with a corrupted graph, often too broken to traverse and make sense of, or apply tools to, is often effectively impossible.  But because the simulation is simply an object graph it is easy to copy the entire simulation. Therefore the simulator has been extended to by default take a copy of itself on every scavenge or old space reclamation, and to run the collection in the copy, performing a leak check afterwards (if enabled).  Hence if a bug occurs or validation fails it will fail in the copy while the original remains as is, yet to be corrupted.  We then take a fresh copy and use it to debug, repeating the process as many times as we like, an approach we call Lemming Debugging.  One throws the copy off the cliff and if it gets to the bottom without incident it is descarded and the simulation continues in the original.  This approach was key in rapid development of the new memory manager that supports both 32 and 64 bits \cite{Forwarders}.

@@

In this section we first discuss the VM simulator's limitations and how we work around them, and then the interaction between the simulator and bootstrapping processes. Lastly, we mention some relevant related work.

@@

%However, so far, most production VMs (Java, Javascript, etc.), even after the huge recent investments in the Javascript VMs by the four major web vendors, are still written in C/C++ and are not metacircular. 
%Hence, although a metacircular VM has interesting advantages, it is not clear it is that convenient to build a VM in such a way.
